---
title: "Assessing predictions"
author: "Ciaran Evans"
output: beamer_presentation
---

## Last time: assessing binary predictions

**Thresholding:** $\widehat{Y}_i = \begin{cases} 1 & \widehat{p}_i \geq c \\ 0 & \widehat{p}_i < c \end{cases}$

**Confusion matrix:**

|  | $Y = 0$ | $Y = 1$ |
| --- | --- | --- |
| $\widehat{Y} = 0$ | True Negatives (TN)  | False Negatives (FN) |
| $\widehat{Y} = 1$ | False Positives (FP) | True Positives (TP)  |

\vspace{2cm}

## Last time


```{r, echo=F, message=F, warning=F}
library(glmhelpR)
m_dengue <- glm(Dengue ~ Age + WBC + PLT, 
                family = binomial, data = dengue)
table(predicted = ifelse(m_dengue$fitted.values > 0.33,
                         "positive", "negative"),
      actual = ifelse(dengue$Dengue == 1, "positive", "negative")) |>
  knitr::kable() |>
  kableExtra::add_header_above(c("Predicted Status" = 1, "Actual Status" = 2)) |>
  kableExtra::kable_styling(bootstrap_options = c("bordered", "hover"))
```


```{r, echo=F}
m_dengue <- glm(Dengue ~ Age + WBC + PLT, 
                family = binomial, data = dengue)

table(predicted = ifelse(m_dengue$fitted.values > 0.2,
                         "positive", "negative"),
      actual = ifelse(dengue$Dengue == 1, "positive", "negative")) |>
  knitr::kable() |>
  kableExtra::add_header_above(c("Predicted Status" = 1, 
                                 "Actual Status" = 2)) |>
  kableExtra::kable_styling(bootstrap_options = c("bordered", "hover"))
```

```{r, echo=F}
m_dengue <- glm(Dengue ~ Age + WBC + PLT, 
                family = binomial, data = dengue)

table(predicted = ifelse(m_dengue$fitted.values > 0.5,
                         "positive", "negative"),
      actual = ifelse(dengue$Dengue == 1, "positive", "negative")) |>
  knitr::kable() |>
  kableExtra::add_header_above(c("Predicted Status" = 1, 
                                 "Actual Status" = 2)) |>
  kableExtra::kable_styling(bootstrap_options = c("bordered", "hover"))
```


## Changing thresholds

## ROC Curves

```{r, echo=F, message=F, warning=F, fig.width=6, fig.height=3}
library(tidyverse)
m_dengue <- glm(Dengue ~ Age + WBC + PLT, 
                data = dengue, family = binomial)

roc_obj <- pROC::roc(m_dengue$y, 
                     m_dengue$fitted.values)

m_dengue_small <- glm(Dengue ~ Age, 
                data = dengue, family = binomial)

roc_obj_small <- pROC::roc(m_dengue_small$y, 
                           m_dengue_small$fitted.values)

data.frame(tpr = c(roc_obj$sensitivities, 
                   roc_obj_small$sensitivities), 
           fpr = c(1 - roc_obj$specificities,
                   1 - roc_obj_small$specificities),
           model = c(rep("Age, WBC, PLT", 
                         length(roc_obj$sensitivities)),
                     rep("Age only", 
                         length(roc_obj_small$sensitivities)))) |>
  ggplot(aes(x = fpr, y = tpr, color = model)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  theme_bw() +
  labs(x = "False Positive Rate (1 - Specificity)", 
       y = "True Positive Rate (Sensitivity)")
```

\vspace{2cm}

## Another example: loan defaults

Data on 948 real estate loans from CA, which were guaranteed by the US Small Business Administration (SBA). Variables include:

* `Default`: whether the business defaulted on the loan (1 = yes, 0 = no)
* `UrbanRural`: 1 if business is in urban area, 2 if business is in rural area, 0 if unknown
* `NewExist`: 1 if business already existed, 2 if business is new, 0 if unknown
* `Term`: Length of the loan term (months)
* `NoEmp`: Number of employees of the business before receiving the loan
* `GrAppv`: Gross amount of loan approved by the bank, in dollars
* `SBA_Appv`: Amount of the loan guaranteed by the SBA, in dollars
* `DisbursementGross`: The amount of money disbursed (loaned), in dollars

## Another example: loan default

Fit a model on the 948 loans:

```{r, echo=F}
m_sba <- glm(Default ~ log(GrAppv) + log(NoEmp + 1) + 
                Term, 
              data = sba_ca_train, family = binomial)

summary(m_sba)$coefficients |>
  round(digits = 3) |>
  knitr::kable()
```

```{r, echo=F}
table(predicted = ifelse(m_sba$fitted.values > 0.5,
                         "yes", "no"),
      actual = ifelse(m_sba$y == 1, "yes", "no")) |>
  knitr::kable() |>
  kableExtra::add_header_above(c("Default? (predicted)" = 1, 
                                 "Default? (actual)" = 2)) |>
  kableExtra::kable_styling(bootstrap_options = c("bordered", 
                                                  "hover"))
```



## Predicting on new data?

\begin{small}

Predictions on the original 948 loans used to fit the model:


```{r, echo=F}
table(predicted = ifelse(m_sba$fitted.values > 0.5,
                         "yes", "no"),
      actual = ifelse(m_sba$y == 1, "yes", "no")) |>
  knitr::kable() |>
  kableExtra::add_header_above(c("Default? (predicted)" = 1, 
                                 "Default? (actual)" = 2)) |>
  kableExtra::kable_styling(bootstrap_options = c("bordered", 
                                                  "hover"))
```
Acc. = 0.816, Sens. = 0.804, Spec. = 0.826, Deviance = 836.06


Predictions on a \textbf{new} set of 407 loans (not used to fit model):

```{r, echo=F}
pred_probs <- predict(m_sba, 
                      newdata = sba_ca_test, 
                      type="response")

table(predicted = ifelse(pred_probs > 0.5,
                         "yes", "no"),
      actual = ifelse(sba_ca_test$Default == 1, "yes", "no")) |>
  knitr::kable() |>
  kableExtra::add_header_above(c("Default? (predicted)" = 1, "Default? (actual)" = 2)) |>
  kableExtra::kable_styling(bootstrap_options = c("bordered", "hover"))
```
Acc. = 0.762, Sens. = 0.732, Spec. = 0.790, Deviance = 405.78
\end{small}

## Predicting on new data

**In general:** A model predicts **better** on the data used to fit the model than it does on **new** data.

**Why?**

\vspace{4cm}

## Class activity

Work on the class activity (course website)

\vspace{4cm}

## Terminology and notation

## Loss functions

## Risk (error)








