---
title: "Model diagnostics"
author: "Ciaran Evans"
output: beamer_presentation
---

## Warmup 

Work on the warmup activity (handout), then we will discuss as a class.

\vspace{4cm}

## Warmup

```{r, include=F}
library(tidyverse)
library(GGally)
library(patchwork)
scorecard <- read_csv("https://sta112-s26.github.io/projects/scorecard.csv")
```

```{r, echo=F, fig.width=4, fig.height=3.5}
p1 <- scorecard |>
  mutate(repayment = ifelse(RPY_3YR_70 == 0, "no", "yes")) |>
  ggplot(aes(x = repayment,
             y = UGDS)) +
  geom_boxplot(size = 0.4, outlier.size = 0.5) +
  labs(x = "At least 70% of students repaying loans?",
       y = "Number of undergrads") +
  facet_wrap(~CONTROL) +
  theme_bw() +
  theme(text = element_text(size = 5))

p2 <- scorecard |>
  mutate(repayment = ifelse(RPY_3YR_70 == 0, "no", "yes")) |>
  ggplot(aes(x = repayment,
             y = NPT4)) +
  geom_boxplot(size = 0.4, outlier.size = 0.5) +
  labs(x = "At least 70% of students repaying loans?",
       y = "Annual cost (USD)") +
  facet_wrap(~CONTROL) +
  theme_bw() +
  theme(text = element_text(size = 5))

p1/p2
```


## Warmup

```{r, echo=F, fig.width=3.5, fig.height=3.5}
scorecard %>%
  ggpairs(columns = 4:10,
          upper = list(continuous = wrap("points", size = 0.5)),
          lower = list(continuous = GGally::wrap(ggally_cor, stars = F))) +
  theme_classic() +
  theme(text = element_text(size = 5))
```

## Outliers and multicollinearity

**Outliers:** Observations which don't follow the same trend as the rest of the data.

\vspace{2cm}

**Multicollinearity:** (linear) relationship between 2 or more explanatory variables.

\vspace{2cm}

Both can cause problems for estimating parameters in a regression model.

## Example: outliers

```{r, echo=F, warning=F, fig.width=5, fig.height=3.5}
library(robustbase)
foodstamp |>
  ggplot(aes(x = ifelse(participation == 0, "no", "yes"), 
             y = log(income + 1))) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Participation in food stamp program", 
       y = "log(Income + 1)")
```

## Example: outliers

With potential outlier:

```{r}
#| echo: FALSE
m1 <- glm(participation ~ log(income + 1), family = binomial,
          data = foodstamp)

summary(m1)$coefficients |>
  round(digits = 3) |>
  knitr::kable()
```

Without potential outlier:

```{r}
#| echo: FALSE
m2 <- glm(participation ~ log(income + 1), family = binomial,
          data = foodstamp[-5,])

summary(m2)$coefficients |>
  round(digits = 3) |>
  knitr::kable()
```

## Cook's distance

## Handling outliers

Should I just remove any unusual observations in the data?

\vspace{4cm}

## Example: multicollinearity

```{r, echo=F, fig.width=3.5, fig.height=3.5}
scorecard %>%
  ggpairs(columns = 4:10,
          upper = list(continuous = wrap("points", size = 0.5)),
          lower = list(continuous = GGally::wrap(ggally_cor, stars = F))) +
  theme_classic() +
  theme(text = element_text(size = 5))
```

## Example: multicollinearity

With all the test score variables:

```{r}
#| echo: FALSE

m1 <- glm(RPY_3YR_70 ~ SATVRMID + SATMTMID + SATWRMID +
            ACTCMMID + ACTENMID + ACTMTMID + ACTWRMID,
          family = binomial, data = scorecard)

summary(m1)$coefficients |>
  round(digits = 3) |>
  knitr::kable()
```

With just one test score variable:

```{r}
#| echo: FALSE

m1 <- glm(RPY_3YR_70 ~ SATVRMID,
          family = binomial, data = scorecard)

summary(m1)$coefficients |>
  round(digits = 3) |>
  knitr::kable()
```

## Variance inflation factors

## Handling multicollinearity

```{r, include=F}
scorecard <- scorecard |>
  select(-c(INSTNM, STABBR, RPY_3YR_RT_SUPP, MD_EARN_WNE_P10,
            GRAD_DEBT_MDN_SUPP))
```


```{r}
m1 <- glm(RPY_3YR_70 ~ .,
          family = binomial, data = scorecard)

car::vif(m1)
```

What should we do when there is high multicollinearity?

\vspace{2cm}



## Summary

* Outliers:
    * Change our coefficient estimates
    * Identify with EDA before model fitting
    * Also check with Cook's distance after model fitting
    * Clear errors, or points from a different population of interest, can be removed. Otherwise, report results with and without outliers
    
\vspace{0.5cm}

* Multicollinearity:
    * Makes it hard to interpret meaning of a coefficient (how do we hold other variables constant??)
    * Increases variability of estimated coefficients
    * Identify with pairs plots, correlation matrices before model fitting
    * Also check with VIFs after model fitting
    * Options: use a subset of the explanatory variables; combine several explanatory variables (average, PCA, etc.); ignore (if only predicting)








