---
title: "Model selection"
author: "Ciaran Evans"
output: beamer_presentation
---

## Last time: Activity

* LRT will choose the full model if $G > \chi^2_{q, \alpha}$
* AIC will choose the full model if $G > 2q$
* LRT is more strict when $\chi^2_{q, \alpha} > 2q$

```{r}
alpha = 0.05
q <- 1:10
qchisq(1 - alpha, q) - 2*q
```
```{r}
alpha = 0.01
q <- 1:20
qchisq(1 - alpha, q) - 2*q
```

## AIC vs. LRT

* LRT can only be used with nested models; AIC can compare non-nested models
* LRT is more stringent when $q$ (difference in number of parameters between models) is small
* AIC is more stringent when $q$ is large
* But, they ask different questions!
    * LRT: Is difference in model deviance due to chance?
    * AIC: Estimate expected prediction error
* **Neither** is asking whether difference in models is **meaningful** (*practical significance*)

## Automated model search

**Potential use case:** 

* *Many* explanatory variables
* Research question is not an inferential question about a particular variable of interest
* Want to identify a subset of potentially useful variables for predicting response

**Some types:**

\vspace{3cm}

## Best subset selection (BSS)

* Fit *all* possible models (for every possible combination of the explanatory variables)
* Choose the model with the best error

\vspace{4cm}

## Stepwise procedures

**Forward selection:**

\vspace{2cm}

**Backward selection:**

\vspace{2cm}

**Bidirectional selection:**

\vspace{1cm}

## Example{.small}

```{r, include=F}
library(tidyverse)
library(glmhelpR)

dengue_sub <- dengue |>
  select(Sex, Age, Vomiting, Abdominal, Temperature,
         BMI, WBC, HCT, PLT, Dengue) |>
  mutate(Sex = as.factor(Sex))
```

```{r}
dengue_bss <- bestglm::bestglm(dengue_sub, binomial, "AIC")

dengue_bss$BestModel$coefficients
```

## Example{.small}

```{r}
# forward selection, beginning with empty model
m_null <- glm(Dengue ~ 1, data = dengue_sub, family = binomial)
m_forward <- MASS::stepAIC(m_null, scope = ~ Sex + Age + 
                             Vomiting + Abdominal + Temperature + 
                             BMI + WBC + HCT + PLT, 
                           direction = "forward", 
                           trace = 0)

m_forward$coefficients
```

## Example{.small}

```{r}
# backward selection, beginning with full model
m_full <- glm(Dengue ~ Sex + Age + Vomiting + Abdominal + 
                Temperature + BMI + WBC + HCT + PLT, 
              data = dengue_sub, family = binomial)
m_backward <- MASS::stepAIC(m_full, 
                            direction = "backward", 
                            trace = 0)

m_backward$coefficients
```

## Example

Simulate data with $n = 100$ observations and 15 explanatory variables, but no population relationship between $\mathbf{x}$ and $Y$.

```{r, include=F}
set.seed(2)
n <- 100
d <- 15
X <- matrix(rnorm(n*(d-1)), nrow = n)
y <- rbinom(n, 1, 0.5)
```

Backward selection (AIC):

```{r, echo=F}
m_full <- glm(y ~ .,
          data = data.frame(X, y), family = binomial)
backward_aic <- MASS::stepAIC(m_full,
                      direction = "backward",
                      trace = 0)
backward_aic$coefficients
```

Forward selection (AIC):

```{r, echo=F}
m_null <- glm(y ~ 1, data = data.frame(X, y), family = binomial)
forward_aic <- MASS::stepAIC(m_null, scope = formula(m_full), 
                             direction = "forward",
                             trace = 0)
forward_aic$coefficients
```
## Example

Forward selection (BIC):

```{r, echo=F}
m_null <- glm(y ~ 1, data = data.frame(X, y), family = binomial)
forward_aic <- MASS::stepAIC(m_null, scope = formula(m_full), 
                             direction = "forward",
                             trace = 0, k = log(n))
forward_aic$coefficients
```

\vspace{4cm}

## Key points

* Search methods are trying to find some set of explanatory variables that does reasonably well at predicting the response
* BSS considers all possible models (may be computationally expensive!)
* Stepwise procedures are greedy alternatives
* Different stepwise procedures don't always agree
* Using BIC tends to give a smaller model than AIC

## Activity

Work on the activity (handout), then we will discuss.


## Activity{.small}

```{r eval=F}
# start with only an intercept
initial_fit <- glm(y ~ 1, data = train_data, family = binomial)

# add predictors one at a time based on AIC
selected_fit <- step(initial_fit,
                     scope = formula(glm(y ~ ., 
                                         data = train_data,
                                         family = binomial)),
                     direction = "forward", trace = 0,
                     steps = 25)

# 10-fold cross validation estimate of prediction error
k <- 10
boot::cv.glm(train_data, selected_fit, 
             cost = logistic_deviance, # loss function to use
             K = k)$delta[1] * k/nrow(train_data) # average deviance
```

## Cross-validation and model selection

**Crucial:** We must cross-validate the whole model selection process, not just the final chosen model!

\vspace{6cm}

## Activity

Work on the activity (course website).













