---
title: "Model selection"
author: "Ciaran Evans"
output: beamer_presentation
---

## Last time: Cross-validation and model selection

**Crucial:** We must cross-validate the whole model selection process, not just the final chosen model!

\vspace{6cm}

## Today

* Model selection and hypothesis testing: can I perform valid inference after choosing a model? (activity and discussion)

\vspace{1cm}

* Cross-validation and model selection: activity

\vspace{3cm}

## Activity 1: Model selection and hypothesis testing

Work on the first class activity (course website)

## Activity 1

Distribution of reported p-values across all variables:

```{r, echo=F, fig.width = 4, fig.height = 2.5}
set.seed(7)

n <- 500
d <- 100
x <- matrix(rnorm(n*d),nrow=n)
y <- rbinom(n, 1, 0.5)
df <- data.frame(y=y,x)

m1 <- glm(y ~ ., data = df, family = binomial)
hist(summary(m1)$coefficients[,4], 
     xlab = "p-values", main = "Histogram of p-values")

pvals <- summary(m1)$coefficients[,4]
```

```{r}
mean(pvals < 0.05)
```


## Activity 1

Distribution of reported p-values for only the *selected* variables:

```{r, echo=F, fig.width = 4, fig.height = 2.5}
set.seed(7)

n <- 500
d <- 100
x <- matrix(rnorm(n*d),nrow=n)
y <- rbinom(n, 1, 0.5)
df <- data.frame(y=y,x)

m1 <- glm(y ~ ., data = df, family = binomial)

m_null <- glm(y ~ 1, data = df, family = binomial)
m_forward <- MASS::stepAIC(m_null, scope = list(upper = m1), 
                           direction = "forward", 
                           trace = 0)

hist(summary(m_forward)$coefficients[,4], 
     xlab = "p-values", main = "Histogram of p-values",
     xlim = c(0, 1))

pvals <- summary(m_forward)$coefficients[,4]
```

```{r}
mean(pvals < 0.05)
```

**Question:** Why is testing hypotheses on the selected variables problematic?

## The problem

* Model selection will choose the variables that appear related to the response *in the observed data*
* If we test only the selected variables, and use the same data for selection and testing, then we have biased the outcome ahead of time: we *know* that these variables must appear related to $Y$ in the data!
* As a result, we lose type I error control

\vspace{3cm}

## Data splitting

## Limitations of automated search methods

* Not suitable if we have particular inferential questions to test
* BSS can be computationally expensive
* Stepwise procedures are greedy and are not guaranteed to find an optimal model (or even a "good" model)
* Automated procedures **don't** do any of the following:
    * Check model diagnostics
    * Experiment with different transformations
    * Assess model performance
    * Identify unusual observations
    * etc.



## Activity

Work on the activity (course website).













