---
title: "Poisson regression"
author: "Ciaran Evans"
output: beamer_presentation
---

## Count variables

**Data:** Data on medical facilities and doctors from a sample of 53 different counties in the US. Variables include:

* `MDs`: the number of medical doctors in the county
* `Hospitals`: the number of hospitals in the county

**Research question:** Can we model the relationship between the number of hospitals and the number of doctors?

\vspace{2cm}


## Plotting the data

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=3.5, fig.height=2.5}
library(tidyverse)
library(Stat2Data)
data("CountyHealth")

CountyHealth %>%
  ggplot(aes(x = Hospitals, y = MDs)) +
  geom_point(size = 0.5) +
  labs(x = "Number of hospitals",
       y = "Number of doctors") +
  theme_bw()
```

**Question:** Does a linear regression model seem appropriate for this relationship?

\vspace{2cm}

## Trying a transformation

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=3.5, fig.height=2.5}

CountyHealth %>%
  ggplot(aes(x = Hospitals, y = sqrt(MDs))) +
  geom_point(size = 0.5) +
  labs(x = "Number of hospitals",
       y = "sqrt(Number of doctors)") +
  theme_bw() 
```

Is a linear regression model appropriate now?

\vspace{2cm}

## Limitation of transformations


## Poisson regression


## Fitting the Poisson regression model

```{r}
m1 <- glm(MDs ~ Hospitals, data = CountyHealth, 
          family = poisson)
summary(m1)$coefficients
```

\vspace{4cm}


## Interpreting the Poisson regression model

```{r}
m1 <- glm(MDs ~ Hospitals, data = CountyHealth, 
          family = poisson)
summary(m1)$coefficients
```

\vspace{4cm}

## Exponential dispersion models

## Activity

Work on the activity (handout).













