---
title: "Measuring error"
author: "Ciaran Evans"
output: beamer_presentation
---


## Take-aways from last class activity

* Even if the model is correctly specified, prediction performance on the **training** data is a biased estimate of prediction performance on **new** data

\vspace{1cm}

* The bias gets worse as the number of explanatory variables in the model increases!

\vspace{1cm}


## Bias of the training error

Let $\mathcal{T} = \{(\mathbf{x}_1, Y_1),...,(\mathbf{x}_n, Y_n)\}$ be a training set, and let $\widehat{h}$ be a model fit on $\mathcal{T}$ by maximizing the likelihood.

Two big results (under regularity conditions):

\vspace{6cm}

## Deriving bias of the training error

Let $\mathbf{y} = (Y_1,...,Y_n) \overset{iid}{\sim} g$ for some unknown distribution $g$ be our **training set**, and let $\mathbf{y}^* = (Y_1^*,...,Y_n^*) \overset{iid}{\sim} g$ be another sample from $g$, *independent* of $\mathbf{y}$.

\vspace{6cm}

## Error quantities

Let $\mathbf{y} = (Y_1,...,Y_n) \overset{iid}{\sim} g$ for some unknown distribution $g$ be our **training set**, and let $\mathbf{y}^* = (Y_1^*,...,Y_n^*) \overset{iid}{\sim} g$ be another sample from $g$, *independent* of $\mathbf{y}$.

\vspace{6cm}

## Idea

$$
\begin{aligned}
\mathcal{R}_{\text{train}}(\widehat{\boldsymbol{\theta}}) &= -\frac{2}{n}\log f_{\widehat{\boldsymbol{\theta}}}(\mathbf{y}) \\
\mathcal{R}_{\mathcal{T}}(\widehat{\boldsymbol{\theta}}) &= \mathbb{E}_{\mathbf{y}^*}\left[ -\frac{2}{n}\log f_{\widehat{\boldsymbol{\theta}}}(\mathbf{y}^*) \bigg| \mathbf{y}  \right] \\
\mathcal{R} &= \mathbb{E}_{\mathbf{y}}[\mathcal{R}_{\mathcal{T}}(\widehat{\boldsymbol{\theta}})] = \mathbb{E}_{\mathbf{y}, \mathbf{y}^*}\left[-\frac{2}{n}\log f_{\widehat{\boldsymbol{\theta}}}(\mathbf{y}^*)\right]
\end{aligned}
$$

\vspace{5cm}

## Lemma 1 proof

## AIC

Let $\widehat{h}$ be a model fit on training data $\mathcal{T}$ by maximizing likelihood (equivalently, minimizing deviance). Let $p$ be the number of parameters in the model (i.e. for logistic regression, $\boldsymbol{\beta} \in \mathbb{R}^p$).

\vspace{5cm}

## Example

```{r, echo=F, message=F}
library(glmhelpR)
```

```{r}
m_dengue <- glm(Dengue ~ Age + WBC + PLT, 
                data = dengue, family = binomial)
m_dengue$deviance
m_dengue$aic
```

\vspace{2cm}







