---
title: "Model selection"
author: "Ciaran Evans"
output: beamer_presentation
---


## Warmup

Work on the warmup activity (handout).

\vspace{4cm}

## Warmup

Which model would you prefer?

```{r include=F}
library(glmhelpR)
```

```{r}
m_dengue1 <- glm(Dengue ~ Age + WBC + PLT, 
                      data = dengue, 
                      family = binomial)
m_dengue1$deviance
```

```{r}
m_dengue2 <- glm(Dengue ~ Age + WBC + PLT + HCT,
                 data = dengue, family = binomial)
m_dengue2$deviance
```

## Another example

```{r}
m_dengue1 <- glm(Dengue ~ Age + WBC + PLT, 
                      data = dengue, 
                      family = binomial)
m_dengue1$aic
```

```{r}
m_dengue2 <- glm(Dengue ~ Age + WBC + PLT +
                        BMI + Temperature + Vomiting,
                 data = dengue, family = binomial)
m_dengue2$aic
```

## Another example

```{r echo=F, fig.width = 5, fig.height=4, warning=F, message=F}
library(tidyverse)
library(patchwork)

# function for computing deviance for binary logistic regression
logistic_deviance <- function(y, phat){
  -2*sum(dbinom(y, 1, phat, log = T))
}

set.seed(3126)
k <- 10
# create the folds (shuffle in case row order matters)
folds <- sample(ceiling((1:nrow(dengue)) * k/nrow(dengue)),
                nrow(dengue), replace=F)

# compute k-fold CV predictions
cv_preds1 <- rep(NA, nrow(dengue))
cv_preds2 <- rep(NA, nrow(dengue))
for(j in 1:k){
  # hold out fold j
  fold_ids <- which(folds == j)
  dengue_train <- dengue[-fold_ids,]
  dengue_test <- dengue[fold_ids,]
  
  # train on remaining k-1 folds
  m_dengue_cv1 <- glm(Dengue ~ Age + WBC + PLT, 
                      data = dengue_train, 
                      family = binomial)
  m_dengue_cv2 <- glm(Dengue ~ Age + WBC + PLT +
                        BMI + Temperature + Vomiting, 
                      data = dengue_train, 
                      family = binomial)
  
  # predict on heldout fold
  cv_preds1[fold_ids] <- predict(m_dengue_cv1,
                                dengue_test,
                                type = "response")
  cv_preds2[fold_ids] <- predict(m_dengue_cv2,
                                dengue_test,
                                type = "response")
}

# CV error estimate
#logistic_deviance(dengue$Dengue, cv_preds1)/nrow(dengue)
#logistic_deviance(dengue$Dengue, cv_preds2)/nrow(dengue)

roc_obj1 <- pROC::roc(dengue$Dengue, 
                     cv_preds1)
roc_obj2 <- pROC::roc(dengue$Dengue, 
                     cv_preds2)
p1 <- data.frame(tpr = c(roc_obj1$sensitivities,
                   roc_obj2$sensitivities), 
           fpr = c(1 - roc_obj1$specificities,
                   1 - roc_obj2$specificities),
           model = c(rep("Model 1", 
                         length(roc_obj1$sensitivities)),
                     rep("Model 2", 
                         length(roc_obj2$sensitivities)))) |>
  ggplot(aes(x = fpr, y = tpr, color = model)) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  theme_bw() +
  labs(x = "FPR", 
       y = "TPR",
       color = "") +
  theme(legend.position = "bottom")

p2 <- calibration_plot(cv_preds1, dengue$Dengue) +
  labs(title = "Model 1")
p3 <- calibration_plot(cv_preds2, dengue$Dengue) +
  labs(title = "Model 2")

(p1 + p2)/(p3 + plot_spacer())
```

## Key points

* Deviance (on the training data) can **never** increase when I add additional explanatory variables -- don't use raw deviance to compare models with different numbers of predictors
* AIC is a better method to compare models (estimate of predictive ability on new data)
* But, a difference in AIC does not always have **practical significance**
* Other criteria for choosing a model: type of research question, model interpretability, model parsimony, ease of use by practitioners, model diagnostics

## Types of research question

For what different types of research question might I use a regression model?

\vspace{6cm}

## Other ways of comparing models

Alternatives to AIC:

\vspace{2cm}

LRT:

\vspace{2cm}

## Activity

Work on the class activity (handout).

\vspace{4cm}

## Activity

LRT:
$$G = \text{deviance}_{\text{reduced}} - \text{deviance}_{\text{full}}$$
Reject when $G > \chi^2_{q, \alpha}$.

\vspace{5cm}

## Activity

* LRT will choose the full model if $G > \chi^2_{q, \alpha}$
* AIC will choose the full model if $G > 2q$
* LRT is more strict when $\chi^2_{q, \alpha} > 2q$

```{r}
alpha = 0.05
q <- 1:10
qchisq(1 - alpha, q) - 2*q
```
```{r}
alpha = 0.01
q <- 1:20
qchisq(1 - alpha, q) - 2*q
```

## AIC vs. LRT

* LRT can only be used with nested models; AIC can compare non-nested models
* LRT is more stringent when $q$ (difference in number of parameters between models) is small
* AIC is more stringent when $q$ is large
* But, they ask different questions!
    * LRT: Is difference in model deviance due to chance?
    * AIC: Estimate expected prediction error
* **Neither** is asking whether difference in models is **meaningful** (*practical significance*)











