---
title: "Poisson regression"
author: "Ciaran Evans"
output: beamer_presentation
---

## Warmup

Work on the warmup activity (handout)

\vspace{4cm}

## Warmup

```{r, include=F}
library(foreign)
library(tidyverse)
articles <- read.dta("http://www.stata-press.com/data/lf2/couart2.dta")

m_art <- glm(art ~ phd + ment + kid5, 
             family = poisson, data = articles)
#summary(m_art)
```

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept)  0.276493   0.087294   3.167  0.00154 ** 
phd          0.008071   0.026358   0.306  0.75944    
ment         0.026537   0.002001  13.264  < 2e-16 ***
kid5        -0.099696   0.035248  -2.828  0.00468 ** 
```

How would we interpret the coefficient on the number of articles published by the student's mentor?

\vspace{3cm}

## Warmup

```
             Estimate Std. Error z value Pr(>|z|)    
(Intercept)  0.276493   0.087294   3.167  0.00154 ** 
phd          0.008071   0.026358   0.306  0.75944    
ment         0.026537   0.002001  13.264  < 2e-16 ***
kid5        -0.099696   0.035248  -2.828  0.00468 ** 
```

What is the predicted number of articles for a student at a PhD program with a prestige score of 4, whose mentor published 30 articles in the last 3 years, and who has no children?

\vspace{3cm}


## EDMs and GLMs

We say that a distribution belongs to the **exponential dispersion model** family if the distribution's probability function can be written in the form

$$
\begin{aligned}
f(y; \theta, \phi) &= a(y, \phi) \exp \left\lbrace \frac{y \theta - \kappa(\theta)}{\phi} \right\rbrace \\
f(y;\mu, \phi)&= b(y, \phi) \exp \left\lbrace -\frac{d(y, \mu)}{2\phi} \right\rbrace
\end{aligned}
$$


For iid observations $(\mathbf{x}_1, Y_1),...,(\mathbf{x}_n, Y_n)$, a **generalized linear model** (GLM) consists of two components:

$$
\begin{aligned}
Y_i | \mathbf{x}_i &\sim \text{EDM}(\mu_i, \phi) \\
g(\mu_i) &= \mathbf{x}_i^T \boldsymbol{\beta}
\end{aligned}
$$

\vspace{1cm}

## What we know so far

GLMs:

* Estimating $\boldsymbol{\beta}$ (and $\phi$ if needed)
* Key properties of EDMs

Linear and logistic regression:

* Model assessment
* Predictions and prediction error
* Inference
* Model selection

Today: Model assumptions and diagnostics for *Poisson* models

## Poisson regression

$$
\begin{aligned}
Y_i | \mathbf{x}_i &\sim Poisson(\mu_i) \\
\log(\mu_i) &= \mathbf{x}_i^T \boldsymbol{\beta}
\end{aligned}
$$

**Question:** What assumptions is this model making? How would we assess those assumptions?

\vspace{6cm}

## Activity

Use the provided packet to answer the following questions:

* Does the Poisson regression model appear to be a good fit?
* If not, which piece of the model do you think is mis-specified?

\vspace{4cm}

## Rootograms: an overall assessment of model fit

```{r, fig.width=4, fig.height=3}
topmodels::rootogram(m_art)
```

## Rootograms: an overall assessment of model fit

* horizontal axis: possible values for the response variable
* vertical axis (square root scale): predicted count for the model (points) and *observed* counts from the data (bars)
* Assessing whether predicted counts match observed counts
* Lack of fit could be due to:
    * Wrong choice of distribution
    * Additional explanatory variables need to be added to model
    * Issues with shape assumption
    
\vspace{1cm}

**Question:** We didn't make a rootogram for binary logistic regression. Why not?

## (Randomized) quantile residual plots

```{r echo=F, fig.width=4, fig.height=2.5, message=F, warning=F}
articles |>
  mutate(res = statmod::qresid(m_art)) |>
  ggplot(aes(x = phd, y = res)) +
  geom_point(size=0.5) +
  geom_smooth() +
  theme_bw() +
  labs(x = "Prestige of PhD program",
       y = "Quantile residuals")
```

What I am looking for in a quantile residual plot?

\vspace{2cm}

## Examining influential points

**Cook's distance:**

\vspace{3cm}

```{r}
articles[which(cooks.distance(m_art) > 0.1),
         c("art", "kid5", "phd", "ment")]
```


## (Randomized) quantile residual plots

```{r echo=F, fig.width=4, fig.height=2.5, message=F, warning=F}
articles |>
  mutate(res = statmod::qresid(m_art)) |>
  ggplot(aes(x = ment, y = res)) +
  geom_point(size=0.5) +
  geom_smooth() +
  theme_bw() +
  labs(x = "Number of articles published by mentor",
       y = "Quantile residuals")
```

How could I decide whether deviations in the quantile residual plot are just due to random chance?

\vspace{1cm}

## Plot lineup

```{r echo=F, message=F, warning=F, fig.height=3.5, fig.width=4}
library(regressinator)

set.seed(70)
qresid_fn <- function(f){
  data.frame(resid = statmod::qresid(f),
             x = model.frame(f)$ment)
}

model_lineup(m_art, fn = qresid_fn, nsim = 20) |> 
  ggplot(aes(x = x, y = resid)) +
  geom_point(size = 0.4) +
  geom_smooth() +
  facet_wrap(vars(.sample)) +
  labs(x = "Mentor articles", y = "Quantile residuals") +
  theme_bw() +
  theme(
  strip.background = element_blank(),
  strip.text.x = element_blank()
)
```

## Quantile residual plot

Plot without the two rows with the most number of mentor publications:

```{r echo=F, fig.width=4, fig.height=2.5, message=F, warning=F}
articles |>
  mutate(res = statmod::qresid(m_art)) |>
  filter(ment < 60) |>
  ggplot(aes(x = ment, y = res)) +
  geom_point(size=0.5) +
  geom_smooth() +
  theme_bw() +
  labs(x = "Number of articles published by mentor",
       y = "Quantile residuals")
```

## QQ plot

**If** the model is a good fit to the data, randomized quantile residuals should be approximately $N(0, 1)$.

```{r echo=F, message=F, warning=F, fig.height=3, fig.width=3}
articles |>
  mutate(res = statmod::qresid(m_art)) |>
  ggplot(aes(sample = res)) +
  geom_qq(size = 0.5) +
  geom_qq_line() +
  theme_bw() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles")
```

## QQ plot

**If** the model is a good fit to the data, randomized quantile residuals should be approximately $N(0, 1)$.

```{r echo=F, message=F, warning=F, fig.height=3, fig.width=3}
articles |>
  mutate(res = statmod::qresid(m_art)) |>
  ggplot(aes(sample = res)) +
  geom_qq(size = 0.5) +
  geom_qq_line() +
  geom_abline(slope = 1, intercept = 0, 
              lty = 2, color = "orange") +
  theme_bw() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles")
```

## Summary so far

For the Poisson regression on the articles data, here is my perspective:

* There are several potential outliers
* The rootogram shows our predicted counts are somewhat low in the tails of the distribution, compared to the observed counts
* Quantile residual plots suggest the shape assumption is overall pretty reasonable
* The QQ plot suggests that the quantile residuals are approximately normal, but variability of the residuals is higher than expected

**Question:** If this particular model is mis-specified, what do you think is the cause? How could we modify the model?













