---
title: "Inference"
author: "Ciaran Evans"
output: beamer_presentation
---

## Class activity

Work on the class activity, then we will discuss.

\vspace{4cm}

## Class activity

**Question 1:** Is there evidence that keeping a pet bird is associated with a greater risk of lung cancer, after accounting for other lifestyle and environmental factors?

\vspace{4cm}

## Class activity

**Question 1:** Is there evidence that keeping a pet bird is associated with a greater risk of lung cancer, after accounting for other lifestyle and environmental factors?

```{r, include=F, message=F, warning=F}
library(tidyverse)

# import the data
birdkp <- read_csv("https://sta712-f26.github.io/class_activities/birdkp.csv")

# fit logistic regression model
m_bird <- glm(Cancer ~ Bird + Sex + SocEc + Age + 
                SmokeYrs + SmokeRate, 
              data = birdkp, family = binomial)
```

```{r, echo=F}
summary(m_bird)$coefficients
```

\vspace{4cm}

## Class activity

**Question 2:** If an individual does keep a pet bird, is there evidence that smoking further increases their risk of lung cancer, after accounting for other factors?

\vspace{6cm}

## Class activity

**Question 2:** If an individual does keep a pet bird, is there evidence that smoking further increases their risk of lung cancer, after accounting for other factors?

```{r, echo=F}
summary(m_bird)$coefficients
```

\vspace{6cm}

## Class activity

**Question 2:** If an individual does keep a pet bird, is there evidence that smoking further increases their risk of lung cancer, after accounting for other factors?

```{r}
m_full <- glm(Cancer ~ Bird + Sex + SocEc + Age + 
                SmokeYrs + SmokeRate, 
              data = birdkp, family = binomial)

m_reduced <- glm(Cancer ~ Bird + Sex + SocEc + Age,
                 data = birdkp, family = binomial)

m_reduced$deviance - m_full$deviance
```

```{r}
pchisq(m_reduced$deviance - m_full$deviance, df = 2, 
       lower.tail = F)
```

\vspace{2cm}

## Class activity

**Question 3:** Consider two individuals. Individual A is male, a non-smoker, of low socioeconomic status, aged 50, who keeps a bird. Individual B is male, of low socioeconomic status, aged 50, who does not keep a bird, and who has smoked for 20 years with an average of 5 cigarettes per day. Estimate the odds ratio of lung cancer for individual A vs. individual B.

```{r, echo=F}
summary(m_bird)$coefficients
```

\vspace{4cm}

## Class activity

**Question 4:** Get a 95% confidence interval for the odds ratio.

\vspace{6cm}

## Class activity

**Question 4:** Get a 95% confidence interval for the odds ratio.

```{r}
beta_hat <- coef(m_bird)
V <- vcov(m_bird)
C <- matrix(c(0, 1, 0, 0, 0, -20, -5), nrow = 1)
log_or_est <- C %*% beta_hat
log_or_sd <- sqrt(C %*% V %*% t(C))
exp(c(log_or_est - 1.96*log_or_sd, 
      log_or_est + 1.96*log_or_sd))
```

\vspace{3cm}





