---
title: "Activity: Logistic regression"
format: pdf
editor: source
---

\vspace{-1.5cm}

## Data

Researchers collected data on 147 individuals in the Netherlands, 67 of whom kept a bird. The data include the following variables:

* `Cancer`: whether the individual developed lung cancer (0 = no, 1 = yes)
* `Bird`: whether the individual kept a bird at home for at least 6 consecutive months from 5 to 14 years before diagnosis (cases) or examination (controls)
* `Sex`: male or female
* `SocEc`: a rough measure of socioeconomic status. `Low` or `High`, depending on the principal wage earner in the household
* `Age`: age in years
* `SmokeYrs`: years of smoking prior to diagnosis or examination
* `SmokeRate`: average number of cigarettes smoked per day

## Logistic regression model

Using the data, the researchers fit a logistic regression model:

```{r, include=F, message=F, warning=F}
library(tidyverse)

birdkp = read.table(
  "http://www.stat.uchicago.edu/~yibi/s226/birdkeeping.txt",
  header=TRUE)
birdkp <- birdkp |>
  mutate(LC = ifelse(LC == "NoCancer", 0, 1),
         BK = ifelse(BK == "NoBird", 0, 1)) |>
  rename(Cancer = LC,
         Sex = FM,
         SocEc = SS,
         Age = AG,
         SmokeYrs = YR,
         SmokeRate = CD,
         Bird = BK)
```

```{r}
m_bird <- glm(Cancer ~ Bird + Sex + SocEc + Age + 
                SmokeYrs + SmokeRate, 
              data = birdkp, family = binomial)

coef(m_bird)
```

## Questions

1. Interpret the estimated coefficient 1.363 in context.

\vspace{5cm}

2. The estimated coefficient on Age is negative. Why might that be the case? Is there anything you would change in the model?

\vspace{5cm}

3. What is the predicted probability of lung cancer for a male individual who keeps a bird, of "high" socioeconomic status, aged 35, who has never smoked?






