---
title: "Cross validation"
format: html
editor: source
---

## Data

Researchers collected data on 147 individuals in the Netherlands, 67 of whom kept a bird. The data include the following variables:

* `Cancer`: whether the individual developed lung cancer (0 = no, 1 = yes)
* `Bird`: whether the individual kept a bird at home for at least 6 consecutive months from 5 to 14 years before diagnosis (cases) or examination (controls)
* `Sex`: male or female
* `SocEc`: a rough measure of socioeconomic status. `Low` or `High`, depending on the principal wage earner in the household
* `Age`: age in years
* `SmokeYrs`: years of smoking prior to diagnosis or examination
* `SmokeRate`: average number of cigarettes smoked per day

The code below imports the data and fits a logistic regression model:

```{r, message=F, warning=F}
library(tidyverse)

# import the data
birdkp <- read_csv("https://sta712-f26.github.io/class_activities/birdkp.csv")

# fit logistic regression model
m_bird <- glm(Cancer ~ Bird + Sex + SocEc + Age + 
                SmokeYrs + SmokeRate, 
              data = birdkp, family = binomial)
```


## k-fold cross validation

In **k-fold cross validation**, we divide the available data $\mathcal{T}$ into $k$ disjoint subsets (called **folds**) $\mathcal{T}_1,...,\mathcal{T}_k$.

For each $j = 1,...,k$, we train the model on $\mathcal{T} \backslash\mathcal{T}_j$; let $\widehat{h}_{-j}$ denote this resulting model. This gives $k$ different models $\widehat{h}_{-1},...,\widehat{h}_{-k}$. The cross validation estimate of error is then
$$
R_{\text{CV}}(\widehat{h}) = \frac{1}{n} \sum \limits_{i=1}^n L(Y_i, \widehat{h}_{-j(i)}(\mathbf{x}_i))
$${#eq-cv_error_estimate}
where $j(i)$ denotes the fold to which observation $i$ belongs: $(\mathbf{x}_i, Y_i) \in \mathcal{T}_{j(i)}$.

Common values of $k$ are 5 and 10.

## Questions

Here we will implement 5-fold cross validation for the bird data and the logistic regression model fit above.

1. To begin, we need to generate folds. The code below is one way to randomly assign each observation in the data to one of five different folds:

```{r}
k <- 5
# create the folds (shuffle in case row order matters)
folds <- sample(ceiling((1:nrow(birdkp)) * k/nrow(birdkp)),
                nrow(birdkp), replace=F)
```

2. Now let's make a vector to store the cross-validated predictions for each observation in the `birdkp` data. Right now, it will be full of `NA`s, since we haven't filled in the predicted probabilities yet.

```{r}
cv_preds <- rep(NA, nrow(birdkp))
```


2. Divide the `birdkp` data into two pieces: the observations belonging to fold 1 (i.e., where `folds == 1`), and the observations belonging to folds 2--5.

3. Re-fit the logistic regression model on the data in folds 2--5.

4. Using the `predict` function, calculate the predicted probabilities for the observations in fold 1, from the fitted model built on folds 2--5.

5. Fill in the entries in `cv_preds` for fold 1 with the predicted probabilities from question 4.

6. Repeat for each of the other folds in turn (you can make this efficient by using a `for` loop).

7. Compute the deviance with the 5-fold CV predicted probabilities. The following function will be helpful:

```{r}
# function for computing deviance for binary logistic regression
# y = true response values
# phat = predicted probabilities
logistic_deviance <- function(y, phat){
  -2*sum(dbinom(y, 1, phat, log = T))
}
```


```{r, include=F, eval=F}
logistic_deviance <- function(y, phat){
  -2*sum(dbinom(y, 1, phat, log = T))
}

set.seed(3126)
k <- 5
# create the folds (shuffle in case row order matters)
folds <- sample(ceiling((1:nrow(birdkp)) * k/nrow(birdkp)),
                nrow(birdkp), replace=F)

# compute k-fold CV predictions
cv_preds <- rep(NA, nrow(birdkp))
for(j in 1:k){
  # hold out fold j
  fold_ids <- which(folds == j)
  birdkp_train <- birdkp[-fold_ids,]
  birdkp_test <- birdkp[fold_ids,]
  
  # train on remaining k-1 folds
  m_bird_cv <- glm(Cancer ~ Bird + Sex + SocEc + Age + 
                SmokeYrs + SmokeRate, 
              data = birdkp_train, family = binomial)
  
  # predict on heldout fold
  cv_preds[fold_ids] <- predict(m_bird_cv,
                                birdkp_test,
                                type = "response")
}

# CV error estimate
logistic_deviance(birdkp$Cancer, cv_preds)/nrow(birdkp)
```




