---
title: "Warmup: Cross validation"
format: pdf
editor: source
---

\vspace{-1.5cm}


## Model building and cross validation

Your friend collects training data of size $n = 500$ observations, on 1000 different explanatory variables.

They have too many explanatory variables available, so they use forward stepwise selection (with AIC) to choose a model, limiting the number of variables allowed in the model to 25.

In the end, they have a regression model with 25 explanatory variables. Let's call this their `selected model`. To assess the predictive ability of their `selected model`, they use cross validation in the following way:

* Divide the their training data into 10 folds
* Fit the `selected model` (with the same explanatory variables as those chosen by stepwise selection) to 9 of the 10 folds, and predict on the remaining fold
* Repeat, and average across the 10 folds

Here is their code, and the resulting estimate of prediction error:

```{r, include=F}
logistic_deviance <- function(y, phat){
  -2*sum(dbinom(y, 1, phat, log = T))
}

n_train <- 500
n_test <- 1000

# d = number of explanatory variables 
# So d + 1 is the length of beta (including intercept)
d <- 1000

set.seed(47234)
x_train <- matrix(rnorm(n_train*d), nrow=n_train)
y_train <- rbinom(n_train, 1, 0.5)
train_data <- data.frame(x_train, y = y_train)

# generate test data
x_test <- matrix(rnorm(n_test*d), nrow=n_test)
y_test <- rbinom(n_test, 1, 0.5)
test_data <- data.frame(x_test, y = y_test)
```


```{r, warning=F, message=F}
# start with only an intercept
initial_fit <- glm(y ~ 1, data = train_data, family = binomial)

# add predictors one at a time based on AIC
selected_fit <- step(initial_fit,
                     scope = formula(glm(y ~ ., 
                                         data = train_data,
                                         family = binomial)),
                     direction = "forward", trace = 0,
                     steps = 25)

# 10-fold cross validation estimate of prediction error
k <- 10
boot::cv.glm(train_data, selected_fit, 
             cost = logistic_deviance, # loss function to use
             K = k)$delta[1] * k/nrow(train_data) # average deviance
```

Later, they collect a separate test sample, independent of the training sample, which is from the same population. Here is the average deviance calculated on this new test data:

```{r}
# predicted probabilities on test data
pred_probs <- predict(selected_fit, newdata = test_data,
                      type = "response")

# estimate of test error using test data
logistic_deviance(test_data$y, pred_probs)/n_test
```


**Question:** Their 10-fold CV estimate of prediction error is *not* representative of the true prediction error on new data. Why? What should they do differently?

*Hint:* The code is a faithful implementation of their proposed process. The problem is with the process itself.


