---
title: "Activity: saddlepoint approximation and dispersion estimates"
format: html
editor: source
callout-icon: false
---

## Gamma distribution

In homework, you are working on showing that the gamma distribution is an example of an EDM. Written in dispersion model form, the gamma distribution has pdf

$$f(y; \mu, \phi) = \frac{1}{y(\phi e)^{1/\phi} \ \Gamma(1/\phi)} \exp\left\lbrace -\frac{d(y, \mu)}{2\phi} \right\rbrace$$

where the unit deviance is

$$d(y, \mu) = 2\left(-\log\left(\frac{y}{\mu} \right) + \frac{y - \mu}{\mu} \right)$$

### Saddlepoint approximation

Since $V(\mu) = \mu^2$ for the gamma, the saddlepoint approximation is

$$f(y; \mu, \phi) \approx \frac{1}{y\sqrt{2\pi \phi}} \exp\left\lbrace -\frac{d(y, \mu)}{2\phi} \right\rbrace$$

In other words, the saddlepoint approximation uses

$$\frac{1}{y(\phi e)^{1/\phi} \ \Gamma(1/\phi)} \approx \frac{1}{y\sqrt{2\pi \phi}}$$

### Questions

The code below computes $f(y; \mu, \phi)$ and the saddlepoint approximation to the density for the gamma distribution with $\phi = 0.1$ and $\mu = 5$.

```{r}
phi <- 0.1
mu <- 5
y <- seq(0, 10, 0.1)

# approximate normalizing constant with saddlepoint
saddlepoint_constant <- 1/(y * sqrt(2*pi*phi))

# true normalizing constant
original_constant <- (exp(1) * phi)^(-1/phi)/(y * gamma(1/phi))

# unit deviance for the gamma
unit_deviance <- 2*(-log(y/mu) + (y-mu)/mu)

# densities (saddlepoint approximation and true density)
saddlepoint_dens <- saddlepoint_constant * exp(-unit_deviance/(2*phi))
original_dens <- original_constant * exp(-unit_deviance/(2*phi))
```

1. Plot the two densities -- does the saddlepoint approximation do a good job?

2. Try different values of $\phi$. How does the shape of the gamma distribution change as you change $\phi$? How does the quality of the saddlepoint approximation change as you change $\phi$?


## Estimating $\phi$

Suppose we have $Y_1,...,Y_n \overset{iid}{\sim} Gamma(\mu, \phi)$. (No explanatory variables to consider for now). Let's use $\overline{Y}$ as our sample estimate of $\mu$.

Then, the mean deviance and Pearson estimators of $\phi$ are:

$$
\begin{aligned}
\widehat{\phi}_D &= \frac{\sum \limits_{i=1}^n d(Y_i, \overline{Y})}{n-1} = \frac{\sum \limits_{i=1}^n 2\left(-\log\left(\frac{Y_i}{\overline{Y}} \right) + \frac{Y_i - \overline{Y}}{\overline{Y}} \right)}{n-1} \\
\widehat{\phi}_P &= \frac{1}{n-1} \sum \limits_{i=1}^n \frac{(Y_i - \overline{Y})^2}{(\overline{Y})^2}
\end{aligned}
$$

Here is code to simulate $n = 100$ observations from a gamma with $\mu = 5$ and $\phi = 0.1$:

```{r}
mu = 5
phi = 0.1
n <- 100
y <- rgamma(n, shape = 1/phi, scale = mu*phi)
```

### Questions

3. Compute the two estimators, and compare with the true value of $\phi$.

4. Simulate many samples, and compare the bias and variance of the two estimators.

5. How do the bias and variance change as $\phi$ changes?




